Halving the cost of quantum addition

### 2017-09-12 paper.tex
Polishing based on Dave's initial comments from weekly 1:1

Existing Toffoli constructions use at least 4 T gates per Toffoli gate.
The Cuccaro adder uses $2n + O(1)$ Toffoli gates \citep{Cuccaro2004}.
Correspondingly, existing adder constructions have T-counts of $8n + O(1)$ \citep{Amy2013} and this has been the case for over a decade \citep{Barenco1995, Cuccaro2004}.
The leading factor of 8 has been conjectured to be optimal \citep{AustinDiscussionsAndEmails2017}.

@@

We halve the number of T gates needed to perform an adder by halving the number of T gates needed to perform Toffoli gates that appear in compute/uncompute pairs.
Although we focus on adders in this paper, paired Toffoli gates appear in a wide variety of circuits.
We discuss some specific examples in \autoref{sec:other-applications} (e.g. applying the same rotation $R_Z(\theta)$ to several qubits).

Our paper is divided into four sections.
We motivate the need to optimize T-counts and discuss existing work in \autoref{sec:introduction}.
\autoref{sec:construction} presents our $4n + O(1)$ T-count adder and explains its pieces, including the temporary logical-AND.
In \autoref{sec:other-applications}, we discuss several circuit constructions and estimates improved by our adder and temporary logical-AND constructions.
Finally, \autoref{sec:conclusion} summarizes our contributions and discusses future work.

@@

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &x &&\ctrl{1} &\qw &x  && &&          &&\ctrl{2}&\qw     &\targ    &\qTi &\targ     &\qw &\qw &\qw &\\
      &y &&\ctrl{1} &\qw &y  &&=&&          &&\qw     &\ctrl{1}&\targ    &\qTi &\targ     &\qw &\qw &\qw &\\
      &  &&         &\qw &xy && &&|A\rangle &&\targ   &\targ   &\ctrl{-2}&\qT  &\ctrl{-2} &\qH &\qS &\qw &\\
    }
  }
  \vspace*{0.25cm} \\
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &x  &&\ctrl{1} &\qw &x && &&&\qw &\qw    &\ctrl{1}         &\qw &&&&&&&&&&\\
      &y  &&\ctrl{1} &\qw &y &&=&&&\qw &\qw    &\gate{Z}         &\qw &&&&&&&&&&\\
      &xy &&\qw      &    &  && &&&\qH &\meter &\cw \cwx \bullet &    &&&&&&&&&&\\
    }
  }
  \caption{
	How to compute and uncompute the logical-AND of two qubits.
	The computation circuit has a T-count of 4 and a T-depth of 1.
	Note that the $|A\rangle$ state input contributes to the T-count, because $|A\rangle$ states are the resource used to perform T gates.
	The uncomputation circuit only uses Clifford gates, and so has a T-count of zero.
    \\
    An alternative uncomputation construction is to simply do the reverse of the computation circuit.
    This alternative approach has a net T-count of 2 (because an $|A\rangle$ state is recovered).
    The resulting total T-count of 6 for the entire temporary logical-AND would still be an improvement on existing work, but is inferior to the measure-and-fixup construction shown above.
  }
  \label{fig:temporary-logical-AND}
\end{figure}

Decreasing the T-count of addition reduces the T-count of any construction based on addition.
For example, in \citep{Fowler2012} it is estimated that factoring a 2048-bit number on a surface-code-based quantum computer would take 27 hours and $2 \cdot 10^{12}$ distilled $|A\rangle$ states.
The time estimate is based on Toffolis having a T-depth of 3, and the $|A\rangle$ count estimate is based on Toffolis having a T-count of 7.
Because Shor's algorithm is dominated by the cost of controlled additions, our techniques (and previous work) reduce the average T-count and T-depth of its Toffolis to $\approx 2.7$ and $1$ respectively.
This reduces the estimates to 9 hours and $8 \cdot 10^{11}$ distilled $|A\rangle$ states.

@@

For example, our improved adder allows the operation $R_Z(\theta)$ to be applied to $n$ qubits with a total T-cost $4n + O(\text{poly}(\lg \frac{1}{\epsilon}) \lg n)$ as follows.
First, iteratively apply our temporary adder to combine the target qubits into a single binary register storing their pop count (the number that are in the state $|1\rangle$).
This takes $4n + O(\lg n)$ T gates to do.
Then, for each position $k$ in the pop count register, synthesize and apply the operation $R_Z(\theta \cdot 2^k)$ to the register qubit at that position.
This uses a number of T gates proportional to some polynomial of the desired bit precision, times the number of bits in the output register.
The output register is exponentially smaller than $n$, so this cost is negligible for large $n$.
Finally, uncompute the temporary pop count (using no T gates).
This applies the $n$ desired $R_Z(\theta)$ operations.

Another rotation-based operation that can be implemented via an adder is the $n$-bit phase gradient operation $\text{Grad}_n = \sum_{k=0}^{2^n-1} e^{2 i \pi k / 2^n} |k\rangle \langle k|$.
Normally this operation would be implemented by separately applying the operation $R_z(\pi 2^{-p})$ to each qubit of the target register, where $p$ is the qubit's index in the register and the number of T gates needed for each rotation depends on the maximum per-gate error $\epsilon$.
However, assuming a ``phase gradient register" prepared in the state $2^{-b/2} \sum_{k=0}^{2^b-1} e^{2 i \pi k / 2^b} |k\rangle$ is available, the phase gradient operation can be performed via addition using only $4n + O(1)$ T gates.
Add the target register into the phase gradient register, and phase kickback will apply the $\text{Grad}_n$ operation to the target.
Ignoring the one-time cost of initializing the reusable phase gradient register, this uses $4n + O(1)$ T gates.

@@

It is interesting to consider if addition or temporary logical-AND gates can be done with still fewer T gates.
\autoref{fig:lower-bound-logical-AND} presents a simple proof that the T-count of the temporary logical-AND is at least 2.
Regardless of whether the true lower bound is 4, we expect that there are special cases with lower average T-counts.
For example, consider a register being conditionally incremented several times.
Incrementing is performed with paired Toffoli gates that can be replaced by temporary logical-ANDs.
An alternative way to reduce the T-count is by cancelling computation work at the start of an increment against uncomputation work from the end of the previous increment.
The alternative approach happens to have the same T-count in this case, but that may not always be the case.

@@

Regardless of whether there will be further T-count improvements for operations as basic as the temporary logical-AND, it is clear that many other T-count optimizations are waiting to be found.
Not just in the construction of basic low-level operations, but in medium-level constructions, high-level constructions, lower-than-circuit level constructions, and generally across the whole technology stack that will be needed to perform error corrected quantum computation with the surface code for the first time.

>>>>

The Cuccaro adder uses $2n + O(1)$ Toffoli gates \citep{Cuccaro2004}.
Existing constructions implement the Cuccaro adder's Toffolis using $8n + O(1)$ T gates \citep{Barenco1995, Cuccaro2004, Amy2013}.
This T-count can be achieved either by applying the ancilla-and-fixup construction from \citep{Jones2013} to each individual Toffoli gate, or by noting that all but one of the adder's Toffoli gates appear in compute/uncompute pairs and then applying the matched-phase-error construction from \citep{Barenco1995, Nielsen2009} to each pair.

The leading factor of 8 in the T-count of an $n$-bit adder has stood for over a decade \citep{Barenco1995, Cuccaro2004}, and has been conjectured to be optimal \cite{AustinDiscussionsAndEmails2017}.
We disprove this conjecture with a construction based on the temporary logical-AND, which combines ideas from both the matched-phase-error construction and the ancilla-and-fix-up construction.
The temporary logical-AND halves the number of T gates needed to perform Toffoli gates that appear in compute/uncompute pairs, and thereby halves the T-count of the Cuccaro adder from $8n + O(1)$ to $4n + O(1)$.

@@

Our paper is divided into four sections.
We motivate the need to optimize T-counts and discuss existing work in \autoref{sec:introduction}.
\autoref{sec:construction} presents our $4n + O(1)$ T-count adder and explains its pieces, including the temporary logical-AND.
In \autoref{sec:other-applications}, we justify the assertion that our constructions represent widely applicable optimizations by giving examples of quantum tasks and cost projections that are improved by efficient adders and temporary logical-ANDs.
Finally, \autoref{sec:conclusion} summarizes our contributions and discusses future work.

@@

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &x &&\ctrl{1} &\qw &x  && &&          &&\ctrl{2}&\qw     &\targ    &\qTi &\targ     &\qw &\qw &\qw &\\
      &y &&\ctrl{1} &\qw &y  &&=&&          &&\qw     &\ctrl{1}&\targ    &\qTi &\targ     &\qw &\qw &\qw &\\
      &  &&         &\qw &xy && &&|A\rangle &&\targ   &\targ   &\ctrl{-2}&\qT  &\ctrl{-2} &\qH &\qS &\qw &\\
    }
  }
  \vspace*{0.25cm} \\
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &x  &&\ctrl{1} &\qw &x && &&&\qw &\qw    &\ctrl{1}         &\qw &&&&&&&&&&\\
      &y  &&\ctrl{1} &\qw &y &&=&&&\qw &\qw    &\gate{Z}         &\qw &&&&&&&&&&\\
      &xy &&\qw      &    &  && &&&\qH &\meter &\cw \cwx \bullet &    &&&&&&&&&&\\
    }
  }
  \caption{
	How to compute and uncompute the logical-AND of two qubits.
	The computation circuit (top) has a T-count of 4 and a T-depth of 1.
	Note that the $|A\rangle$ state input contributes to the T-count, because $|A\rangle$ states are the resource used to perform T gates.
	The uncomputation circuit (bottom) uses a measure-and-fixup approach \cite{Jones2013} that requires only Clifford gates, and so has a T-count of zero.
    \\
    An alternative uncomputation construction is to simply do the reverse of the computation circuit.
    This alternative approach has a net T-count of 2 (because an $|A\rangle$ state is recovered).
    The resulting temporary logical-AND with a T-count of 6 would still be an improvement on existing work, but would be inferior to the measure-and-fixup approach shown above.
  }
  \label{fig:temporary-logical-AND}
\end{figure}

Decreasing the T-count of addition reduces the T-count of any construction based on addition.
For example, in \citep{Fowler2012} it is estimated that factoring a 2048-bit number on a surface-code-based quantum computer would take 27 hours and $2 \cdot 10^{12}$ distilled $|A\rangle$ states.
The time estimate is based on Toffolis having a T-depth of 3, and the $|A\rangle$ count estimate is based on Toffolis having a T-count of 7.
Because Shor's algorithm is dominated by the cost of controlled additions, our techniques (and previous work) reduce the average T-count and T-depth of its Toffolis to $\sim 2.7$ and $1$ respectively.
This reduces the estimates to 9 hours and $8 \cdot 10^{11}$ distilled $|A\rangle$ states.

@@

For example, our improved adder allows the operation $R_Z(\theta)$ to be applied to $n$ qubits with a total T-cost $4n + O(\text{poly}(\lg \frac{1}{\epsilon}) \lg n)$ as follows.
First, iteratively apply our temporary adder to combine the target qubits into a single binary register storing their Hamming weight.
This takes $4n + O(\lg n)$ T gates to do.
Then, for each position $k$ in the Hamming weight register, synthesize and apply the operation $R_Z(\theta \cdot 2^k)$ to the register qubit at that position.
This uses a number of T gates proportional to some polynomial of the desired bit precision, times the number of bits in the output register.
The output register is exponentially smaller than $n$, so this cost is negligible for large $n$.
Finally, uncompute the Hamming weight register (using no T gates).
This applies the $n$ desired $R_Z(\theta)$ operations.

Another rotation-based operation that can be implemented via an adder is the $n$-bit phase gradient operation $\text{Grad}_n = \sum_{k=0}^{2^n-1} e^{2 i \pi k / 2^n} |k\rangle \langle k|$.
Normally this operation would be implemented by separately applying the operation $R_z(\pi 2^{-p})$ to each qubit of the target register, where $p$ is the qubit's index in the register and the number of T gates needed for each rotation depends on the maximum per-gate error $\epsilon$.
However, assuming a ``phase gradient register" prepared in the state $2^{-b/2} \sum_{k=0}^{2^b-1} e^{-2 i \pi k / 2^b} |k\rangle$ is available, the phase gradient operation can be performed via addition.
Add the target register into the phase gradient register, and phase kickback will apply the $\text{Grad}_n$ operation to the target.
Ignoring the one-time cost of initializing the reusable phase gradient register, this uses $4n + O(1)$ T gates.

@@

It is interesting to consider if addition or temporary logical-AND gates can be done with still fewer T gates.
\autoref{fig:lower-bound-logical-AND} presents a simple proof that the T-count of the temporary logical-AND is at least 2.
Regardless of whether the true lower bound is 4, we expect that there are special cases with lower average T-counts.
For example, if a task involves repeating an action several times (e.g. counting by repeatedly conditionally incrementing a register), T gates from Toffolis in uncomputation work at the end of one repetition may cancel against T gates from Toffolis in computation work at the start of the next repetition.

@@

Regardless of whether there will be further T-count improvements for operations as basic as matched Toffoli gates, it is clear that many other T-count optimizations are waiting to be found.
Not just in the construction of basic low-level operations, but in medium-level constructions, high-level constructions, lower-than-circuit level constructions, and generally across the whole technology stack that will be needed to perform error corrected quantum computation with the surface code for the first time.
